---
title: "advanced Plot"
author: "Chao"
date: "March 15, 2019"
output: html_document
---

#1. draw and save figures separately using ggplot and lapply 

read data

'Plotting_excercise.csv' is the data from last session (ggplot2-20190304)

```{r}
testdata<-read.csv('Plotting_excercise.csv',header = T,sep = ';')
```

##1.1. check names of the variables

```{r}
names(testdata)
```

we want to plot scatterplot to compare:

* RespVar1 vs. VarIndep1

* RespVar1 vs. VarIndep2

* RespVar1 vs. VarIndep3

* RespVar2 vs. VarIndep1

* RespVar2 vs. VarIndep2

* RespVar2 vs. VarIndep3

we have also groups, e.g. "Treatment","Exp" ,"Species","FuncGrp" 

##1.2. load package

```{r library}
library(ggplot2,warn.conflicts = F,quietly = T)
library(scales,warn.conflicts = F,quietly = T)
```

##1.3. build a theme, which will be used throughout the script

```{r mytheme}
mytheme<-
  theme_bw()+        
  theme(plot.background=element_blank(),
        plot.title=element_text(lineheight=0.93,face='bold'),
        panel.background=element_blank(),
        panel.grid.minor=element_blank(),
        panel.grid.major=element_blank(),
        panel.border=element_rect(fill=NULL,colour='black',linetype=1),
        axis.line=element_line(colour='black',linetype=1),
        axis.ticks=element_line(colour='black',linetype=1),
        axis.title.y = element_text(face='bold',
                                    colour='black',size=15,vjust=1),
        axis.text.y  = element_text(colour='black',vjust=0.5, size=10,
                                    face = "bold"),
        axis.title.x = element_text(face='bold',
                                    colour='black',size=10,vjust=0.2),
        axis.text.x  = element_text(colour='black',angle=0, face = "bold",
                                    vjust=0.2, size=8),
        legend.title=element_blank(),
        legend.background=element_blank(),
        legend.text =
          element_text(colour="black", size = 12, face = "bold"),
        legend.position='top',
        legend.key=element_blank(),
        legend.direction="horizontal")
```

##1.4. Scatterplot 

###1.4.1 RespVar1 vs. all VarIndep by group of "FuncGrp", named as individual.plot1 

* RespVar1 vs. VarIndep1

* RespVar1 vs. VarIndep2

* RespVar1 vs. VarIndep3

```{r fig.width=3,fig.height=2}
individual.plot1<-
  lapply(names(testdata)[9:11], function(i){
    ggplot(testdata,aes_string('RespVar1',i))+
      geom_point(aes(color=FuncGrp))+
      geom_smooth(aes(group=FuncGrp,color=FuncGrp),
                 method = 'lm',se=F,size=0.5)+
      mytheme})
```

check the result

```{r}
individual.plot1
```


###1.4.2 Save individual.plot1 (if this one doesnot work, put it into comment)

```{r}
invisible(lapply(seq_along(individual.plot1),function(x)
  ggsave(filename=paste0(names(testdata)[9:11][x],".tiff"),plot=individual.plot1[[x]],
         dpi = 600,compression='lzw',height = 3,width = 3)))
```

###1.4.3 all RespVar vs. all VarIndep by group of "FuncGrp", named as individual.plot2 

```{r fig.width=3,fig.height=2}
 individual.plot2<-
  lapply(names(testdata)[7:8],function(i){
    lapply(names(testdata)[9:11], function(j){
      ggplot(testdata,aes_string(i,j))+
        geom_point(aes(color=FuncGrp))+
        geom_smooth(aes(group=FuncGrp,color=FuncGrp),
                   method = 'lm',se=F,size=0.5)+
        scale_color_manual(values = c('blue','orange'))+
        mytheme})
    })
```

check the result

```{r}
individual.plot2
```

###1.4.4 Save individual.plot2  (if this one doesnot work, put it into comment)

```{r}
invisible(
  lapply(seq_along(individual.plot2),function(x){
    lapply(seq_along(individual.plot2[[x]]),function(y){
      ggsave(filename=paste0(names(testdata)[7:8][x],'_',
                             names(testdata)[9:11][y],".tiff"),
             plot=individual.plot2[[x]][[y]],
             dpi = 600,compression='lzw',height = 3,width = 3)})}))
```

###1.4.5 all RespVar vs. all VarIndep by group of 'Species' and "FuncGrp", named as individual.plot3

```{r fig.width=3,fig.height=4}
 individual.plot3<-
  lapply(names(testdata)[7:8],function(i){
    lapply(names(testdata)[9:11], function(j){
      ggplot(testdata,aes_string(i,j))+
        facet_grid(Species~.,scales = 'free')+
        geom_point(aes(color=FuncGrp))+
        geom_smooth(aes(group=FuncGrp,color=FuncGrp),
                    method = 'lm',se=F,size=0.5)+
        scale_color_manual(values = c('blue','orange'))+
        mytheme})
    })
```

check the result

```{r}
individual.plot3
```


###1.4.6 Save individual.plot3 (if this one doesnot work, put it into comment)

```{r}
invisible(
  lapply(seq_along(individual.plot3),function(x){
    lapply(seq_along(individual.plot3[[x]]),function(y){
      ggsave(filename=paste0('Species_',
                             names(testdata)[7:8][x],'_',
                             names(testdata)[9:11][y],
                             ".tiff"),
             plot=individual.plot3[[x]][[y]],
             dpi = 600,compression='lzw',height = 5,width = 3)})}))
```

###1.4.7 all RespVar vs. all VarIndep by group of 'Species' and "FuncGrp", named as individual.plot4

```{r fig.width=10,fig.height=6}
 individual.plot4<-
  lapply(names(testdata)[7:8],function(i){
    lapply(names(testdata)[9:11], function(j){
      ggplot(testdata,aes_string(i,j))+
        facet_grid(Species~Exp,scales = 'free')+
        geom_point(aes(color=FuncGrp))+
        geom_smooth(aes(group=FuncGrp,color=FuncGrp),
                    method = 'lm',se=F,size=0.5)+
        scale_color_manual(values = c('blue','orange'))+
        mytheme})
    })
```

check the result

```{r}
individual.plot4
```

###1.4.8 Save individual.plot4 (if this one doesnot work, put it into comment)

```{r}
invisible(
  lapply(seq_along(individual.plot4),function(x){
    lapply(seq_along(individual.plot4[[x]]),function(y){
      ggsave(filename=paste0('Species_Exp_',
                             names(testdata)[7:8][x],'_',
                             names(testdata)[9:11][y],
                             ".tiff"),
             plot=individual.plot4[[x]][[y]],
             dpi = 600,compression='lzw',height = 6,width = 8)})}))
```

#2. plot correlation matrix


use chart.Correlation() function from 'PerformanceAnalytics'

```{r}
library(PerformanceAnalytics)
chart.Correlation(testdata[,7:11],histogram=F, pch=19,cex=0.2,na.action=na.omit)
```

plot correlation matrix by groups of Treatment and FuncGrp

results should be followed by 

figure 1: treat1*FG1
figure 2: treat1*FG2
figure 3: treat2*FG1
figure 4: treat2*FG2

```{r}
test.subset<-
  lapply(levels(testdata$Treatment),function(i){ 
  lapply(levels(testdata$FuncGrp),function(j){
        chart.Correlation(
          testdata[testdata$Treatment==i&
                   testdata$FuncGrp==j,7:11],
          histogram=F, pch=19,cex=0.2)
        })
      })
```

use rcorr() from 'Hmisc' package for obtaining correlation matrix

```{r}
library(Hmisc)
lrm.resul<-rcorr(as.matrix(testdata[,7:11]))
```

check the structure of lrm.result

```{r}
str(lrm.resul)
```

plot correlation matrix by using corrplot() function from 'corrplot' package

```{r fig.width=10,fig.height=6}
library(corrplot)
# r value
corrplot(lrm.resul$r,method='col',tl.cex = 1, tl.col='black',cl.cex=1,
         #addCoef.col = "black",
         col= colorRampPalette(c("#D6604D", "#F4A582", "#FDDBC7",
        "#FFFFFF", "#D1E5F0", "#92C5DE", "#4393C3"))(10))

```

```{r}
# p value
corrplot(lrm.resul$P,method='col',tl.cex = 1, tl.col='black',cl.cex=1,
         addCoef.col = "black",
         col= colorRampPalette(c("#D6604D", "#F4A582", "#FDDBC7",
        "#FFFFFF", "#D1E5F0", "#92C5DE", "#4393C3"))(10))
```

#3. 3D plot 

```{r}
#read data
correctF<-read.csv('correctF.csv',header = T,sep='')
#convert colunm of 'Date' to date format
correctF$Date<-as.Date(correctF$Date)
#subset, namely, choose data when 680<wavelength<770
correctF<-correctF[which(correctF$wavelength>680&correctF$wavelength<770),]
```

plot 3D figure using plot_ly() function from 'plotly' package

'ggplot2' uses '+' to add functions, while 'plotly' uses %>% to add functions

```{r}
library(plotly)
F.3d1<-correctF %>% group_by(Date) %>% 
  plot_ly(y = ~Date, x = ~wavelength, z = ~corre.F,type='scatter3d',mode='lines',color=I("BLUE"))
```

check the result

```{r}
F.3d1
```

save the result to 'html' format (if this one doesnot work, put it into comment)

```{r}
htmlwidgets::saveWidget(F.3d1, "F3D1.html")
```

optimize the figure by setting the axis format and text font

```{r}
#set text font
f1 <- list(family = "Times New Roman",size = 10,color = "black")
#set axis format
axis.x<- list(titlefont = f1,tickfont = f1,
            gridcolor="rgb(0,0,0)",ticklen = 5,
            gridwidth = 2.5,title='Wavelength (nm)')
axis.y<- list(titlefont = f1,tickfont = f1,
            gridcolor="rgb(0,0,0)",ticklen = 5,
            gridwidth = 2.5,title='Date')
axis.z<- list(titlefont = f1,tickfont = f1,
            gridcolor="rgb(0,0,0)",ticklen = 5,
            gridwidth = 2.5,title='Fluorescence (mW m-2 sr-1 nm-1)')
```

plot 3D figure by adding axis setting

```{r}
F.3d2<-correctF %>% group_by(Date) %>% 
  plot_ly(y = ~Date, x = ~wavelength, z = ~corre.F,type='scatter3d',mode='lines',color=I("BLUE"))%>% 
  layout(scene=list(xaxis = axis.x,
                    yaxis = axis.y,  
                    zaxis = axis.z))
F.3d2
#save the result(if this one doesnot work, put it into comment)
htmlwidgets::saveWidget(F.3d2, "F3D2.html") 
```


#4. contour plot using plot_ly() funciton from 'plotly' package

setting the text font, axis format, and title format of the legend 

```{r include=FALSE}
#set the font
f1 <- list(family = "Times New Roman",size = 16,color = "black")
# sett the axis
axis.x <- list(titlefont = f1,tickfont = f1,showgrid = T,title='Wavelength (nm)',
               autotick = FALSE,  ticks = "outside",  tick0 = 680,  dtick = 20,ticklen = 5)
axis.y <- list(titlefont = f1,tickfont = f1,showgrid = T,title='Date')
# set the legend title format
legend.title<-list(titlefont = f1,title='Fluorescence (mW m-2 sr-1 nm-1)')
```

plot the contour figure 


```{r}
library(plotly)
F.contour1<-correctF %>% 
  plot_ly(y = ~Date, x = ~wavelength, z = ~corre.F,type='contour',#colors = c('grey','black'),
                             contours = list(size=0.2,showlabels = TRUE),  #contours=list(coloring='heatmap')
                             marker = list(colorbar = list(len = 0.1, 
                                      title = legend.title)))%>%
  layout(autosize = F, 
    margin = list(t = 80, r = 30, b = 160, l = 240),
    xaxis=axis.x,
    yaxis=axis.y)
F.contour1
```

save figure (if this one doesnot work, put it into comment)

```{r}
htmlwidgets::saveWidget(F.contour1, "Fcontour1.html")
```

add two vertical lines using add_segments() function

```{r}
F.contour2<-
  F.contour1%>%
  add_segments(x = 690, xend = 690,inherit=FALSE,  showlegend=FALSE,
               line = list(color='orange',dash = 'dash'),
               y=as.Date('2015-03-04'),
               yend=as.Date('2015-07-20'))%>%
  add_segments(x = 740, xend = 740,inherit=FALSE,  showlegend=FALSE,
               line = list(color='orange',dash = 'dash'),
               y=as.Date('2015-03-04'),
               yend=as.Date('2015-07-20'))
F.contour2
```

check and save figure (if this one doesnot work, put it into comment)

```{r}
htmlwidgets::saveWidget(F.contour2, "Fcontour2.html")
```

