---
title: "Rcourse Session-1"
author: "Chao"
date: "25 syyskuuta 2018"
output: html_document
---

###How to read data

```{r}
gpp<-read.csv('data/gppsmeardata_20 160101120000.csv',header = T,sep = ',')
```

###read files one by one

```{r}
ET<-read.csv('data/ET smeardata_20160101120000.csv',header = T,sep = ',')
NEE<-read.csv('data/NEEsmeardata_20160101120000.csv',header = T,sep = ',')
```

###Combine two files into one dataframe

```{r}
gpp.ET<-merge(gpp,ET,by=c("Year","Month","Day","Hour","Minute", "Second"),all = T)
```

###Read all the files (more than two same format files), and Combine them into one dataframe

```{r}
#list all the files we want to read
filenames=list.files(path='Z:/Chao/Rcourse/Session-1/data', pattern = '*.csv',full.names=TRUE)
#read all the files 
myfile<-lapply(filenames, function(x){
  read.csv(x, stringsAsFactors = F, sep=',')
  })
#using 'reduce' function from 'purrr' package
library(purrr)
df.all<-reduce(myfile,function(x,y) {merge(x, y, all = TRUE)})
#check the head of the data
head(df.all)
```

###Convert the columns to datetime object 

```{r}
#combine date to one column
df.all$date<-paste(df.all$Year,df.all$Month,df.all$Day,sep = '-')
#check date
head(df.all$date)
#combine time to one column
df.all$time<-paste(df.all$Hour,df.all$Minute,df.all$Minute,sep = ':')
#check time
head(df.all$time)
#combine date and time to one column
df.all$datetime<-paste(df.all$date,df.all$time)
#check datetime
head(df.all$datetime)
#conver to datetime format using lubridate library
library(lubridate)
df.all$datetime<-ymd_hms(df.all$datetime)
#check datetime
head(df.all$datetime)
```
###Remove the colunms that we do not need

```{r}
#check the name index firstly
names(df)
df<-df.all[,-c(1:6)]
```

###Reorder the colunms by variable index

```{r}
df<-df[,c(7,6,5,1:4)]
```

###Reset columns or variable names

```{r}
#check the name index firstly
names(df)
colnames(df)[4:7]<-c('ET','GPP','NEE','Sensible_Heat')
#check the name 
names(df)
```


###Plot timeseries using 'ggplot2' package

http://www.cookbook-r.com/Graphs/

```{r}
#for ggplot function
library(ggplot2)
ggplot(df,aes(x=datetime))+
  geom_point(aes(y=GPP))
```

###Plot timeseries by changing the datetime format
```{r fig.width=16}
#for date_breaks function
library(scales)
ggplot(df,aes(x=datetime))+
  geom_line(aes(y=GPP),size=1)+
  geom_point(aes(y=GPP),size=1,color='darkgreen')+
  scale_x_datetime(breaks = date_breaks("1 month"), 
                   labels = date_format("%m-%d\n%H:%M", tz="UTC"),
                   limits = c(as.POSIXct("2016-01-01 00:00:00 UTC"),
                              as.POSIXct("2017-01-01 00:00:00 UTC")))
```


###Plot all timeseries 
```{r fig.width=14,fig.height=12,message=F,warning=F}
#format the data 
library(data.table,warn.conflicts = F,quietly = T)
df.melt<-melt(df,id.vars=c('datetime','date','time'))

library(scales,warn.conflicts = F,quietly = T)
p<-ggplot(df.melt,aes(x=datetime,y=value))+
  facet_grid(variable~.,scales = 'free_y',switch = 'y')+
  geom_point(size=1)+
  geom_line(aes(color=variable),size=1)+
  scale_x_datetime(breaks = date_breaks("1 month"), 
                   labels = date_format("%m-%d\n%H:%M", tz="UTC"),
                   limits = c(as.POSIXct("2016-01-01 00:00:00 UTC"),
                              as.POSIXct("2017-01-01 00:00:00 UTC")))
p
suppressWarnings(print(p))

```

###Add more settings to the plot: 
```{r fig.width=14,fig.height=12,message=F,warning=F}
figure1<-p+xlab('Date')+theme_bw()+
  theme(strip.text = element_text(size=25,color='black'),
        strip.placement= 'outside',
        strip.background = element_blank(),
        legend.position = 'top',
        legend.title = element_blank(),
        legend.text = element_text(size = 20,color = 'black',face='bold'),
        legend.spacing.x = unit(0.5,'cm'),
        plot.title = element_text(size=25,color='black',face='bold'),
        axis.text = element_text(size = 20,color = 'black'),
        axis.title.x = element_text(size = 25,color = 'black'),
        axis.title.y =element_blank())
suppressWarnings(print(figure1))
```






###Pair plot

```{r message=F,warning=F}
library(PerformanceAnalytics,warn.conflicts = F,quietly = T)
#Visualization of a Correlation Matrix,default method is "pearson"
chart.Correlation(df[,c(4:7)], histogram=T)
chart.Correlation(df[,c(4:7)], histogram=F)
```

