---
title: "R1-Introduction"
author: "Chao Jon Yann"
date: "9 lokakuuta 2018"
output: pdf_document
---

#1.variables and Operators

A variable is used to store your data or various information that could be
changed.

##1.1.1. Assign a variable using <- , ->or =
This is how we assign a variable using <- , ->or =

```{r}
x<-1
#now we want to show the contents of x using print function 
print(x)
```

```{r}
y=c(12,3)
print(y)
```

```{r}
c('h','i',9)->z
print(z)
```

##1.1.3. Names of the variables
The name of a variable should be unique, and can consist of letters, numbers, underline or dots. The
variable should start with: a letter or a dot (this dot cannot followed by a number), but cannot start
with a number or underline.
For example, can you find which variable name is not correct?

x.1<-1 ; x_2<-2 ;.x<-3 ; 4y<-5 ; *_z<-6; _2<-7, .2a<-8*

Try here:

```{r}

```


##1.2. Data types

In R, there are 4 basic data types that are widely used, namely, a single data type. 

character: ‘a’ 

numeric: 2,0, or 5.6 

integer: 4L (the L means R will store 4 as integer) 

logical: TRUE, FALSE

The data types of R objects can be the variable types. These variables can consist of a dataset.
Try to find the type of above mentioned variables x,y,z using class(), and the structure of x,y,z using str()
The common used structure of the dataset in R include:
vector (basis), factor, matrix,array, data frame and list.

##1.3. Operaters 

#2. Data structures and Control flow

2.1.1. A vector is the first basic data structure.
Vectors are one dimension array that can consist of numberic, character,or logical data. All the elements
inside a vector should be one data type (numberic, character,or logical).
A vector is formed using function: c(), which combines all the arguments into one common type.
For example (1) Create a vector named as a, store letters of b to e into a, (2) print a, (3) check type of a,
and (4) structure of a.

```{r}
#(1)create a character vector: a
a<-c('b','c','d','e')
#(2)print: a
print(a)
```

```{r}
#(3)check type of a
class(a)
```

```{r}
#(4)check structure of a
str(a)
```

Try to (1) create a vector named as a1, store first 10 numbers into a1,(2) print a1, (3) check type of a1, and
(4) structure of a1.
Which type of above mentiond variable z (z<-c(‘h’,‘i’,9))?

```{r}

```

##2.1.2. A factor is created using a vector

All the elements inside this factor will be character. A factor can efficiently store characters, because only
once per character is stored.

```{r}
#create a vector
treatment<-c('control','control','drought','nutrient','drought')
#check the type of treatment
class(treatment)
```

```{r}
#check the structure of treatment
str(treatment)
```

Now let’s create a factor and check the class and structure of this factor. 

Can you see the difference between treatment and treatment.factor?

```{r}
#create a factor
treatment.factor<-factor(treatment)
#check the type
class(treatment.factor)

```

```{r}
#check the structure
str(treatment.factor)
```


Yes, the numbers of these unique character are called levels, which can be examined using levels() function

```{r}
#check levels of treatment.factor
levels(treatment.factor)
```

Try to create a factor named as plantnum.factor which is from a vector plant.number contains 5 elements
c(1,2,3,3,2). How many levels plantnum.factor have, and what are they? Can you see the difference between
plant.number and plantnum.factor?

```{r}

```

##2.1.3. A list can contain mant types of datasets.

A list is ...

```{r}
alist<-list(c('b','c','d','e'),
            c('control','control','drought'),
            c(1:10),
            c(1L,3L),
            88,
            '%&()',
            c(0.3,2.34,6.57,0),
            3+9,
            c('4','9'),
            list(c(1,2,'g')))
print(alist)
```

```{r}
class(alist)
```

```{r}
str(alist)

```


##2.2. Conditions and Loops using vector: if, for, while, repeat

###2.2.1. if, if else, ifelse

if (condition) {statement} or if (condition) statement

```{r}
x<-10
if (x<5) {print('X is smaller than 5')}
if (x>5) print('X is larger than 5')
```

if (condition) { statement1 } else { statement2 }
or
if (condition) { statement1 } else if (condition2){ statement2 } else if (condition3){ statement3 } else if
(condition4){ statement4 } else { statement5 }

```{r}
y<-20
if (y>10) {
print ('y is larger than 10')
} else {
print('y is smaller than 10')
}
```

```{r}
y<-20
if (y<0) {
print ('y is a negative number')
} else if (y==0) {
print('y is 0')
} else if (y>0 & y <10) {
print('y is greater than 0 and smaller than 10')
} else if (y==10) {
print ('y is 10')
} else if (y>10 & y <20){
print ('y is greater than 10 and smaller than 20')
} else {
print ('y is 20')
}

```



#3. Functions and Applying functions
##3.1.
##3.2.
##3.3. 

##3.4. Extral funcitons

car



